---
title: "BLT: Simulated Annealing over Trees"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{BLT: Simulated Annealing over Trees}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The other vignettes estimate one piece of the problem at a time: the mutation
rates with the tree given, or the coalescent times with the topology given.
This vignette runs the whole method, referred to as BLT, which estimates all
three together. It alternates a simulated annealing search over topologies with
refits of the rates and of the times.

```{r setup}
library(blt)
library(ape)
```

## The algorithm

`run_blt()` carries out the following steps.

1. A UPGMA tree on the Hamming distances between the lumped mutation states
   gives the starting topology and the starting intercoalescent times.
2. The mutation rates are estimated from the observed mutation counts and the
   total branch length of that tree.
3. The intercoalescent times are fitted on the starting topology by gradient
   ascent, for at most `max_steps` steps.
4. Then for `max_outer_iter` cycles:
    a. simulated annealing over topologies, `sa_max_iter` proposals, each one
       scored by `inner_sa_steps` steps of gradient ascent on the times;
    b. the rates are refitted from the new times;
    c. the times are refitted on the new topology, for at most
       `inner_max_steps` steps.

The height of the tree is held at the height of the UPGMA tree throughout,
because the likelihood on its own cannot separate the overall scale of the tree
from the overall mutation rate.

## Simulating data

One tree, one integration barcode, four cut sites. As in the other vignettes
the topology comes from `phylodyn` and the alleles from
`simulate_data_finite_alleles()`, but here nothing about the tree is passed to
the estimator: only the alleles at the leaves are observed.

```{r}
set.seed(1)
n <- 6 # number of cells
s <- 4 # number of cut sites

simul1 <- phylodyn::coalsim(samp_times = 0, n_sampled = n, traj = exp_traj,
                            method = "tt", val_upper = 11)
true_tree <- phylodyn::generate_newick((simul1))$newick
true_height <- coalescent.intervals(true_tree)$total.depth

states_matrix <- state_space_matrix(s)
theta_true <- matrix(0.5, nrow = s, ncol = s)
q_true <- q_matrix(states_matrix, theta_true)
m_true <- matrix(20, nrow = s, ncol = s)

d <- simulate_data_finite_alleles(true_tree, q_true, states_matrix,
                                  m_true)[[2]][1:n, ]
rownames(d) <- true_tree$tip.label
d
```

An entry `"0"` is an uncut site, `"3:10"` is allele 10 of a cut at site 3, and
`"14:14"` is allele 14 of a single cut spanning sites 1 through 4. Cells 4, 5
and 6 were each cut across all four sites at once, which wipes out everything
that came before and leaves them looking unrelated to each other even though
they need not be.

## Setting up the search

The starting point is the UPGMA tree, which is what the estimator has to
improve on.

```{r}
lumped_d <- t(apply(d, 1, lumped_state))
dist_matrix <- hamming_dist_matrix(lumped_d)
dimnames(dist_matrix) <- list(rownames(d), rownames(d))
upgma_tree <- phangorn::upgma(as.dist(dist_matrix))
matching(upgma_tree)
```

Topologies are proposed by `adjacent_swap()`, which exchanges a child of one
coalescent event with a child of the next event back in time. Every proposal is
one such swap away from the current tree.

```{r}
set.seed(2)
adjacent_swap(matching(upgma_tree))
```

A proposal is accepted with probability 1 if it raises the log-likelihood, and
otherwise with a probability that falls as the temperature falls, so that early
proposals explore and later ones do not. Under the default `"exp"` schedule
`sa_temp_update()` drops the temperature geometrically by `sa_alpha` after every
proposal.

```{r}
sa_accept_prob(e1 = -40, e2 = -39, temp = 1) # an improvement
sa_accept_prob(e1 = -40, e2 = -45, temp = 1) # a big drop, early on
sa_accept_prob(e1 = -40, e2 = -45, temp = 0.1) # the same drop, later on
```

The mutation rates are refitted from the observed counts and the current times
by `refit_theta()`, using the posterior mean of the Dirichlet-Poisson model in
the *Initial Estimate of Theta* vignette. Because a coalescent tree always has
$n, n-1, \ldots, 2$ lineages in its successive intervals, the total branch
length depends on the times but not on the topology.

```{r}
c_vec <- flatten_upper(observed_mutations(d, s, lumped_d)$obsM)
init_ict <- coalescent.intervals(upgma_tree)$interval.length
refit_theta(c_vec, init_ict, time_obs = true_height,
            alpha_prior = rep(5, s * (s + 1) / 2))$lambda
```

## Running BLT

The settings below are the ones used for sa_v1: an initial gradient ascent of
at most 200 steps, then two cycles, each with 30 simulated annealing proposals
scored by 20 gradient steps apiece, followed by 100 steps to refit the times.
They are also the defaults of `run_blt()`. This takes several minutes.

```{r}
result <- run_blt(d, s = s, time = true_height,
                  eta = 1e-3, eps = 1e-3,
                  max_steps = 200,
                  max_outer_iter = 2,
                  sa_max_iter = 30,
                  inner_sa_steps = 20,
                  inner_max_steps = 100,
                  sa_init_temp = 1, sa_alpha = 0.95, sa_schedule = "exp",
                  verbose = TRUE)
```

## Results

The log-likelihood after the initial fit and after each cycle:

```{r}
result$loglik_trace
result$n_accept
```

The search moves the topology towards the truth. Robinson-Foulds distance to
the tree the data was simulated on, for the UPGMA starting point and for the
BLT estimate:

```{r}
c(upgma = phangorn::RF.dist(result$upgma_tree, true_tree),
  blt = phangorn::RF.dist(result$tree, true_tree))
```

```{r fig.width = 7, fig.height = 3.2, echo = FALSE}
op <- par(mfrow = c(1, 3), mar = c(2, 1, 3, 1))
plot(true_tree, main = "true tree")
plot(result$upgma_tree, main = "UPGMA (start)")
plot(result$tree, main = "BLT")
par(op)
```

Finally the estimated relative mutation rates, against the ones the data was
simulated under. `run_blt()` reports `theta` on an absolute scale, with the
overall rate `lambda` folded in, so both are scaled to sum to one.

```{r}
est <- flatten_upper(result$theta)
truth <- flatten_upper(theta_true)
round(data.frame(true = truth / sum(truth), estimated = est / sum(est)), 4)
result$lambda
```
