---
title: "Estimating Effective Population Size N_e"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{estimating_n_e}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(blt)
```

First simulate the tree using the bounded coalescent model.

```{r}
num_tips <- 100
s <- 3

simul <- coalsim_bounded(
  samp_times = 0,
  n_sampled  = num_tips,
  traj       = blt::exp_traj,
  bound      = 1,
  val_upper  = 11
)
mytree <- phylodyn::generate_newick(simul)$newick
mytree$coal_times <- phylodyn::summarize_phylo(mytree)$coal_times
```

The bounded coalescent model depends on a trajectory that we can specify. To evaluate our method, we test several different trajectories including an exponential trajector, constant trajectory, and a piecewise trajectory.

```{r}
constant_traj <- function(time) {
  rep(5, length(time))
}


expo_traj <- function(time) {
  exp(4 - time)
}

exp_traj2 <- function(time) {
  5 * blt::exp_traj(time)
}
```

To estimate coalescent times, we compare 4 methods. The first method is BNPR.

```{r}
bnpr_plot <- function(tree, traj) {
  fit <- phylodyn::BNPR(mytree)
  bnpr_time_grid <- fit$x
  bnpr_ne_mean <- fit$effpopmean
  bnpr_ne_lo <- fit$effpop025
  bnpr_ne_hi <- fit$effpop975
  bnpr_ne_true <- traj(bnpr_time_grid)

  plot(
    bnpr_time_grid, bnpr_ne_mean,
    type = "n",
    log = "y",
    xlim = rev(range(bnpr_time_grid)),
    ylim = range(c(bnpr_ne_lo, bnpr_ne_hi, bnpr_ne_true), finite = TRUE),
    xlab = "",
    ylab = "",
    cex.axis = 1.5,
    asp = 1
  )
  polygon(
    c(bnpr_time_grid, rev(bnpr_time_grid)),
    c(bnpr_ne_lo, rev(bnpr_ne_hi)),
    col = "grey80",
    border = NA
  )
  lines(bnpr_time_grid, bnpr_ne_mean, lwd = 2)
  lines(bnpr_time_grid, bnpr_ne_true, lty = 2)
}

bnpr_plot(mytree, expo_traj)
```

The second method is using the skyline MLE estimation in the `ape` package, which does not account for the bounded coalescent in the model.

```{r}
ape_skyline_plot <- function(mytree, traj, title = "ape_skyline") {
  sky <- ape::skyline(mytree)
  time_grid <- sky$time
  ne_hat <- sky$population.size
  ne_true <- traj(time_grid)

  plot(
    time_grid, ne_hat,
    type = "n",
    log = "y",
    xlim = rev(range(time_grid)),
    ylim = c(0.01, max(c(ne_hat, ne_true), na.rm = TRUE)),
    xlab = "",
    ylab = "",
    cex.axis = 1.5,
    asp = 1
  )

  lines(time_grid, ne_hat, lwd = 2, type = "s")
  lines(time_grid, ne_true, lty = 2)
}

ape_skyline_plot(mytree, expo_traj)
```

The 3rd method is the skyline MLE estimation using our method which accounts for the bounded coalescent model.

```{r}
skyline_plot <- function(tree, traj, title) {
  tree_height <- ape::coalescent.intervals(mytree)$total.depth
  new_fit <- bounded_skyline_ascent(mytree, bound = tree_height)

  new_time_grid <- new_fit$grid
  new_ne <- new_fit$Ne

  new_ne_true <- traj(new_time_grid)

  plot(
    new_time_grid, new_ne,
    type = "n",
    log = "y",
    xlim = rev(range(new_time_grid)),
    ylim = c(.01, max(c(new_ne, new_ne_true), na.rm = TRUE)),
    xlab = "",
    ylab = "",
    cex.axis = 1.5,
    asp = 1
  )

  lines(new_time_grid, new_ne, lwd = 2, type = "s")
  lines(new_time_grid, new_ne_true, lty = 2)
}
skyline_plot(mytree, expo_traj)
```

The final method is the MLE estmation over a grid of times that also accounts for the bounded coalescent model.

```{r}
mle_plot <- function(tree, traj, title) {
  tree_height <- ape::coalescent.intervals(mytree)$total.depth
  mle_fit <- mle_bounded_coal(mytree, bound = tree_height, lengthout = 50)
  mle_ne <- mle_fit$effpop
  mle_time_grid <- mle_fit$grid

  plot(
    mle_time_grid, mle_ne,
    type = "s",
    log = "y",
    xlim = rev(range(mle_time_grid)),
    ylim = range(c(mle_ne, traj(mle_time_grid)), finite = TRUE),
    xlab = "",
    ylab = "",
    cex.axis = 1.5,
    asp = 1
  )

  lines(mle_time_grid, traj(mle_time_grid), lty = 2)
}
mle_plot(mytree, expo_traj)
```

