---
title: "Estimating Coalescent Times"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Estimating Coalescent Times}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette estimates the coalescent times of a lineage tracing tree. A tree
is simulated, alleles are simulated along it, and then everything is held fixed
except the branch lengths: the topology, the mutation rates and the number of
alleles per site are all given, and only the intercoalescent times are
estimated.

```{r setup}
library(blt)
library(ape)
```

## Simulating the tree

`phylodyn` supplies the tree topology (it can be installed following these
[instructions](https://github.com/JuliaPalacios/phylodyn/tree/master)).

```{r}
set.seed(123)
n <- 5 # number of tips or cells
s <- 3 # number of cut sites

simul1 <- phylodyn::coalsim(samp_times = 0, n_sampled = n, traj = exp_traj,
                            method = "tt", val_upper = 11)
mytree <- phylodyn::generate_newick((simul1))$newick
```

The quantities we want to recover are the intercoalescent times
$u = (u_1, \ldots, u_{n-1})$, where $u_k$ is the length of the interval during
which the tree has $k + 1$ lineages. `ape::coalescent.intervals()` returns them
for the simulated tree, and the total depth is the height of the tree.

```{r}
intervals <- coalescent.intervals(mytree)
true_ict <- intervals$interval.length # intercoalescent times u_1, ..., u_{n-1}
tree_height <- intervals$total.depth
true_ict
```

The topology is held fixed throughout, and is represented by the matching
returned by `matching()`. Row $r$ stores the two children and the parent of the
$r$th coalescent event, counting from the present.

```{r}
pairs <- matching(mytree)
pairs
```

The likelihood is a function of the $2n - 2$ branch lengths
$b = (b_1, \ldots, b_{2n-2})$, but every branch length is a sum of
intercoalescent times, so $b = A u$ where $A$ is the $(2n - 2) \times (n - 1)$
matrix returned by `decompose_branches()`. The optimization below is therefore
carried out over $u$, the $n - 1$ intercoalescent times, rather than over $b$.
$b$ is only ever computed from $u$ as $Au$. This is what keeps the tree
ultrametric, since any $u$ at all gives a tree whose tips are all at the same
height.

```{r}
a <- decompose_branches(pairs)
a
true_b <- a %*% matrix(true_ict, ncol = 1)
```

## The mutation model

Each of the `s` sites is either uncut or carries one of a number of alleles. A
single cut can span several adjacent sites at once, in which case it overwrites
whatever those sites carried before. `state_space()` and `state_space_matrix()`
enumerate the resulting mutation states, and `q_matrix()` turns a matrix of
rates `theta` into the rate matrix `q` over those states, where `theta[i, i]`
is the rate of a cut at site `i` alone and `theta[i, j]` the rate of a single
cut spanning sites `i` through `j`.

The rates are held at the values the data is simulated under, since only the
times are being estimated. `m` is the number of distinct alleles a cut can
produce, and `t_prob` is the matrix of transition probabilities over one unit
of time.

```{r}
states <- state_space(s)
states_matrix <- state_space_matrix(s)
theta <- matrix(1, nrow = s, ncol = s)
q <- q_matrix(states_matrix, theta)
eigen_q <- eigen(q)
eigen_q$inv <- solve(eigen_q$vectors)
t_prob <- eigen_q$vectors %*% diag(exp(eigen_q$values)) %*% eigen_q$inv
m <- matrix(20, nrow = s, ncol = s)
```

## Simulating data on the fixed tree

`simulate_data_finite_alleles()` starts from the unmutated state at the root
and simulates the mutation process down the tree. Only the leaves are observed,
and `haplotypes()` turns the observed alleles into the initial forward
probabilities used by the likelihood.

```{r}
sim <- simulate_data_finite_alleles(mytree, q, states_matrix, m)
d <- sim[[2]][1:n, ] # allele states at the leaves
hap <- haplotypes(d, n, s)
d
```

`branch_loglik()` evaluates the log-likelihood of the data for a given set of
branch lengths. Evaluating it at the true branch lengths gives the value that
the optimizer is trying to reach.

```{r}
true_loglik <- branch_loglik(pairs, d, hap, true_b, eigen_q, states,
                             states_matrix, m, t_prob, approx = 1)
true_loglik
```

## Estimating the times

`estimate_coal_times()` maximizes this log-likelihood over the intercoalescent
times by gradient ascent with an AdaGrad step size. The gradient with respect
to the branch lengths comes from the forward and backward probability vectors,
and the chain rule $\partial \ell / \partial u = A^{T} \partial \ell /
\partial b$ turns it into a gradient with respect to the times. After every
step the times are projected back onto the set $\sum_k u_k \le T$, so the
estimated tree never grows taller than `total_time`. Whenever a step produces a
non-positive time or lowers the log-likelihood it is retried with the learning
rate divided by five; the learning rate is never restored afterwards, so it can
only decrease over the course of a run.

The starting point is a perturbation of a flat set of times, rescaled so that
it already has the right total height.

```{r}
start <- rep(1, n - 1) + runif(n - 1, min = -0.2, max = 0.2)
start <- start * tree_height / sum(start)
start
```

```{r}
result <- estimate_coal_times(pairs, d, hap, ict = start,
                              eta = 0.001, eps = 0.001,
                              q = q, eigen_q = eigen_q,
                              states = states, states_matrix = states_matrix,
                              m = m, t_prob = t_prob,
                              total_time = tree_height, approx = 1,
                              max_steps = 1000)
result$steps
result$logliks[length(result$logliks)]
```

Comparing the estimated intercoalescent times with the truth, and the
coalescent times measured from the present:

```{r}
data.frame(
  start = start,
  true_ict = true_ict,
  est_ict = result$ict,
  true_coal_time = cumsum(true_ict),
  est_coal_time = result$coal_times
)
```

`tree_from_pairs()` puts the estimated times back onto the fixed topology, so
the two trees can be drawn side by side. They differ only in their branch
lengths.

```{r}
est_tree <- tree_from_pairs(pairs, result$coal_times)
est_tree$tip.label <- mytree$tip.label
```

```{r fig.width = 7, fig.height = 3.5, echo = FALSE}
op <- par(mfrow = c(1, 2), mar = c(3, 1, 3, 1))
plot(mytree, main = "true tree", x.lim = c(0, tree_height))
axisPhylo()
plot(est_tree, main = "estimated tree", x.lim = c(0, tree_height))
axisPhylo()
par(op)
```
