---
title: "Initial Estimate of Theta"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Initial Estimate of Theta}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

First load the packages needed. `phylodyn` contains functions to simulate tree topologies (can be installed using these instructions (GitHub)[https://github.com/JuliaPalacios/phylodyn/tree/master]. `blt` contains functions that will simulate the states of each node and help estimate mutation rates.

```{r setup}
library(blt)
library(ape)
```

simulate the tree topology. For these simulations, we will use a fixed tree topology and simulate multiple datasets on it.

```{r}
set.seed(123)
n <- 5 # number tips or cells
s <- 3 # number of cut sites

# the tree is fixed, we are simulating multiple realizations
simul1 <- phylodyn::coalsim(samp_times = 0, n_sampled = n, traj = exp_traj,method="tt",val_upper=11)
mytree <- phylodyn::generate_newick((simul1))$newick
summary <- ape::coalescent.intervals(mytree)
treelength <- sum(summary$lineages*summary$interval.length)
```

simulate data at each node given the tree topology `mytree`.

```{r}
states <- state_space(s)
states_matrix <- state_space_matrix(s)
theta <- matrix(c(1, .5, .5,
                  0,  1, .5,
                  0,  0,  1), nrow=3, byrow=TRUE)
q <- q_matrix(states_matrix, theta)

num_sim <- 50 # number of simulations
datalist <- list()
# d_list <- list()
for (i in 1:num_sim){
  ## simulates and returns two matrices. 1 mutation states, 2 allele states
  m = matrix(20, nrow = s, ncol = s)
  datalist[[i]]= simulate_data_finite_alleles(mytree, q, states_matrix, m) 
  d = datalist[[i]][[2]][1:n,]
  # d_list[[i]] = d
}
```

Method 1 uses the method of moments.

```{r}
t <- coalescent.intervals(mytree)$total.depth
mom_theta_vec_list <- matrix(0, nrow = num_sim, ncol = s*(s+1)/2) # store theta estimates from MoM
for (i in 1:num_sim) {
  lumped_d = datalist[[i]][[1]][1:n,]
  a <- observed_mutations(d, s, lumped_d)
  b <- a$obsM/n

  # Initial value of theta for optimization
  theta0 <- rep(1,(s*(s+1))*.5)
  theta0 <- theta0*0.03

  # probs_function computes the sum of (observed - expected)^2 for each mutation type
  est_theta1 <- optim(theta0, fn = probs_function,  observed_muts = b, t = t, s = s,
                     states = states, states_matrix = states_matrix, lower = rep(0, length(theta0)), method = "L-BFGS-B")

  theta_vec <- est_theta1$par
  mom_theta_vec_list[i,] <- theta_vec
}

# scale theta to sum to 1 for each simulation
scaled_mom_theta_vec_list <- matrix(0, nrow = num_sim, ncol = s*(s+1)/2)
rates_list <- c()
for (i in 1:num_sim) {
  rates_list[i] <- sum(mom_theta_vec_list[i,])
  scaled_mom_theta_vec_list[i,] <- mom_theta_vec_list[i,]/rates_list[i]
}
# boxplot(scaled_mom_theta_vec_list, outline=FALSE) 
```
Method 2 uses the posterior mean of the following model
$$
\begin{align*}
\theta &\sim \text{Dirichlet}(\alpha) \\
\lambda &\sim \text{Exp}(\beta) \\
C_i | \theta_i &\sim \text{Poisson}(\lambda t \theta_i) \\
\end{align*}
$$
$\theta$ is the vector of relative mutation rates, $\lambda$ is the overall mutation rate, $C_i$ is the observed number of mutations of type $i$, and $t$ is the total tree length.

```{r}
## save results for posterior mean method
d_theta <- s*(s+1)/2 # dimension of theta
alpha <- rep(1, d_theta) # prior parameters
beta <- 1 # prior parameter for lambda
post_mean_theta_vec_list <- matrix(0, nrow = num_sim, ncol=d_theta)  
post_mean_lambda_vec_list <- c()

# Methods 2 and 3 require upgma tree estimation, and the package causes conflicts with optim, so it is run separately.

for (i in 1:num_sim){
  d <- datalist[[i]][[2]][1:n,]
  lumped_d <- datalist[[i]][[1]][1:n,]
  a <- observed_mutations(d, s, lumped_d)
  
  # Posterior distribution depends on the total length of branches
  # need to estimate UPGMA tree first
  # UPGMA tree estimation depends on the data, so must be inside the loop, after new data is simulated
  mytree_dist_matrix <- hamming_dist_matrix(lumped_d) ## this should be different every loop
  mytree_upgma <- phangorn::upgma(mytree_dist_matrix)
  upgma_tree_scale <- t/sum(coalescent.intervals(mytree_upgma)$total.depth) ## scale factor to make total depth = t
  l <- sum(coalescent.intervals(mytree_upgma)$lineages*coalescent.intervals(mytree_upgma)$interval.length)*upgma_tree_scale ## sum of all tree branches
  
  ## ----------------------- Method 2: Estimate theta using posterior mean--------------------------
  c <- a$obsM # Counts
  c <- flatten_upper(c) # make C a vector
  post_mean_theta_vec_list[i,] <- (alpha + c)/(sum(c) + sum(alpha))
  post_mean_lambda_vec_list[i] <- (sum(c) + 1)/(beta + l) # beta = 1, l = tree length (exposure time)
}


boxplot(post_mean_theta_vec_list, outline=FALSE) 
```

Method 3 uses the posterior max of the following model.

```{r}
## MAP method
map_theta_vec_list <- matrix(0, nrow = num_sim, ncol=d_theta)
map_lambda_vec_list <- c()

# Methods 2 and 3 require upgma tree estimation, and the package causes conflicts with optim, so it is run separately.

for (i in 1:num_sim){
  d <- datalist[[i]][[2]][1:n,]
  lumped_d <- datalist[[i]][[1]][1:n,]
  a <- observed_mutations(d, s, lumped_d)
  
  # Posterior distribution depends on the total length of branches
  # need to estimate UPGMA tree first
  # UPGMA tree estimation depends on the data, so must be inside the loop, after new data is simulated
  mytree_dist_matrix <- hamming_dist_matrix(lumped_d) ## this should be different every loop
  mytree_upgma <- phangorn::upgma(mytree_dist_matrix)
  upgma_tree_scale <- t/sum(coalescent.intervals(mytree_upgma)$total.depth) ## scale factor to make total depth = t
  l <- sum(coalescent.intervals(mytree_upgma)$lineages*coalescent.intervals(mytree_upgma)$interval.length)*upgma_tree_scale ## sum of all tree branches
  
  ## ----------------------- Method 3: Estimate theta using MAP------------------------------------------
  c <- a$obsM # Counts
  c <- flatten_upper(c) # make c a vector
  map_theta_vec_list[i,] <- (alpha + c - 1)/(sum(c) + sum(alpha) - 1) # d_theta = dimension of theta
  map_lambda_vec_list[i] <- sum(c)/(beta + l) # beta = 1, l = tree length
}

boxplot(map_theta_vec_list, outline=FALSE) 
```
